----
title: "EDE Group Project"
author: "Gaby Czarniak, Mara Michel, Sam Sedar"
date: "2023-11-14"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(here); here()
library(rvest)
library(cowplot)
library(agricolae)
library(dplyr)
library(tidyverse)
library(lubridate)
library(trend)
library(zoo) 
library(Kendall)
library(tseries)

getwd()

```


# Question 3: Is there a relationship between power plant distribution and impacts to human health?

Null hypothesis: there is no relationship between power plant distribution and impacts to human health.

Among the emissions types tracked as part of the eGRID data set used in Questions 1 and 2, nitrogen oxides (NOx) and sulfur dioxide merit further exploration, as they form part of criteria air pollutants regulated by EPA. Under the Clean Air Act, US EPA has set National Ambient Air Quality Standards on six criteria air pollutants: particulate matter (PM2.5 and PM10), atmospheric ozone, carbon monoxide, lead, nitrogen dioxide, and sulfur dioxide. These pollutants reach widespread exposure--millions of people--due to their numerous and diverse sources. Causal or likely causal evidence pointing to a variety of negative health endpoints exists for each one of them. 

Power plants have historically made up a major stationary point source of air pollution, including SO2 and NO2, in particular. For nitrogen dioxide (NO2) and sulfur dioxide (SO2), the regulatory emissions standard-setting has been most heavily informed by research on respiratory effects. It should be noted that nitrogen oxides (NOx) also contribute to the formation of atmospheric ozone, which causes respiratory problems. 

The following section of analyses explore relationships between emissions of SO2 and NOx and respiratory effects using eGRID data as well as data from the Environmental Burden and Health Vulnerability modules in the Environmental Justice Index (EJI). The development of the EJI is relatively new and the index has limitations. In accordance with its limitations, which can be found in detail within the technical documentation linked in the Metadata folder, this high-level exploration is not intended to define individuals at risk, represent risk or exposure for a community, or suggest that the markers explored form a full or near-full representation of environmental or health characteristics. This brief exploration merely examines whether a statistically significant relationship can be detected among select local factors such as power plant emissions and human health effects at a point in time across North Carolina. Based on established knowledge regarding air pollution and its impacts to human health, the expectation would be that higher emissions of NOx and SO2 are associated with higher rates of respiratory disease.

```{r Distribution of Respiratory Illness}

### DEFINITIONS ###

# E_PM = Annual mean days above PM2.5 regulatory standard - 3-year average
# E_DSLPM = Ambient concentrations of diesel PM/m3
# E_OZONE = Annual mean days above O3 regulatory standard - 3-year average
# EP_BPHIGH = Percentage of individuals with Raw high blood pressures values
# EP_ASTHMA = Percentage of individuals with asthma
# PLNOXAN = Plant Annual NOx Emissions (PLNOXAN) – The total annual NOx emissions, in short tons, for the plant. Biogas components are adjusted. For CHP plants, the value is adjusted by the electric allocation factor. This adjusted emissions field is estimated by first making the biogas adjustment (if it exists) and then applying the electric allocation factor (if the plant is a CHP).
# PLSO2AN = Plant Annual SO2 Emissions (PLSO2AN) – The total annual SO2 emissions, in short tons, for the plant. Landfill gas components are adjusted. For CHP plants, the value is adjusted by the electric allocation factor. This adjusted emissions field is estimated by first making the landfill gas adjustment (if it exists) and then applying the electric allocation factor (if the plant is a CHP).
# NOx_emissions = sum of NOx emissions by county across plants
# SO2_emissions = sum of SO2 emissions by county across plants
# NOx_map = map of NOx emissions by county
# NOx_counties_sf_join = attribute of total NOx emissions attached to spatial frame
# SO2_counties_sf_join = attribute of total SO2 emissions attached to spatial frame

#### DATA WRANGLING ####

#Establish shared GEOID format
EJI_countytract$CountyFIPS <- stringr::str_extract(EJI_countytract$GEOID, "^.{5}")

#Get 2021 spatial census data to convert tract GEOID to mapped coordinates
spatial_tracts <- tigris::tracts(state = 37) %>% 
  st_as_sf()
class(spatial_tracts)
st_crs(spatial_tracts) # NAD1983, EPSG 4269

#Join the eGrid attributes to the county spatial features
tracts_EJI_sf_join <-  merge(x = spatial_tracts,
                           y = EJI_countytract, 
                           by.x = 'GEOID', 
                           by.y = 'GEOID' )
#Some EJI tracts did not GEOIDs that matched the spatial tracts

#Select variables of interest
tracts_EJI_sf_select <- tracts_EJI_sf_join %>%
  select(GEOID, CountyFIPS, TRACTCE.x, StateAbbr, COUNTY, E_TOTPOP, M_TOTPOP, 
         E_PM, E_OZONE, E_DSLPM, E_COAL, E_LEAD, EP_UNINSUR, EP_AGE65, 
         EP_ASTHMA, EP_BPHIGH, EP_CANCER, EP_MHLTH, geometry) %>%
  rename(TractCode = TRACTCE.x)

#Map asthma prevalence by tract
Asthma_by_tract_map <- ggplot(data = tracts_EJI_sf_select) +
  geom_sf(data = tracts_EJI_sf_join)+
  geom_sf(aes(fill = EP_ASTHMA)) +
  labs(title = "Estimated Asthma Prevalence by NC Tract",
       subtitle = "2021",
       fill = "Percentage of Individuals with Asthma")
print(tracts_map)
```



```{r Power Plant Data by Tract}

#Check coordinates of plant geometry (point) data
st_crs(plant_gen_sf$geometry) # NAD1983, EPSG 4269

#Match coordinates of plants_NC to tract geometry
plants_within_tracts <- plant_gen_sf[tracts_EJI_sf_select,]
tracts_containing_plants <- tracts_EJI_sf_select[plant_gen_sf,]

#Map of plants across NC tracts, shown over NC tract boundaries 
mapview(spatial_tracts, col.regions='pink', layer.name = 'North Carolina Tracts') + 
 mapview(plants_within_tracts,
          layer.name = 'Plants Within Each Tract',
          alpha=.5)

#Map of just NC tracts that contain power plants
mapview(spatial_tracts, col.regions='grey', layer.name = 'North Carolina Tracts') + 
  mapview(tracts_containing_plants, col.regions=, layer.name = 'NC Tracts Containing Power Plants')

#Zoom in on tracts within one county: Robeson, the county with highest total number of existing plants in 2021
#Filter for Robeson tracts
Robeson_tracts <- tracts_EJI_sf_select %>%
  filter(COUNTY == "Robeson")
#Filter for Robeson plants
tract_plants_Robeson <- filter(plants_within_tracts, CNTYNAME == "Robeson")

#Show plants within Robeson tracts, specifically
mapview(Robeson_tracts, col.regions='pink', 
        layer.name = 'Robeson County Tracts, NC') +
 mapview(tract_plants_Robeson,
          layer.name = 'Plants Within Each Tract',
          alpha=.5)

#Convert NC eGrid tabular county data to spatial data
plants_NC_sf <- Plant_NC %>% st_as_sf(coords = c('LON','LAT'), crs=4269)

#Table of plant data with the tract that corresponds to each plant
intersection_plants_EJI <- st_intersection(plants_NC_sf, tracts_EJI_sf_select)

#Exporting to CSV:
#write.csv(intersect_plants_EJI,'Data/Processed/EBM_Processed.csv', row.names=FALSE)
```


```{r Relationship between Asthma Prevalence and Power Plant Count}
##Visualize the relationship between asthma prevalence and number of power plants per tract ##

#Power plant count
plantcount_per_tract <- intersection_plants_EJI %>%
  group_by(GEOID) %>%
  summarise(plantcount = n())

#Asthma prevalence by tract
HealthImpacts_by_tract <- intersection_plants_EJI %>% 
  select(GEOID, EP_ASTHMA, EP_BPHIGH) %>%
  st_drop_geometry()
class(HealthImpacts_by_tract)

#Join plant count and 
join_HealthImpacts_plantcount <- merge(HealthImpacts_by_tract, plantcount_per_tract, 
                                    by = "GEOID")
#Remove duplicates
join_HealthImpacts_plantcount <- distinct(join_HealthImpacts_plantcount, GEOID, .keep_all = TRUE)

#Check normal distributions of continuous variables
qqnorm(join_HealthImpacts_plantcount$EP_ASTHMA); qqline(join_HealthImpacts_plantcount$EP_ASTHMA)
qqnorm(join_HealthImpacts_plantcount$EP_BPHIGH); qqline(join_HealthImpacts_plantcount$EP_BPHIGH)
#The data does not follow a normal distribution

#Bartlett's test checks null hyp that the variance in each of the groups are the same. We want to run this because ANOVA is robust against departures from equal variance, however, if the test result is false, we can still run ANOVA. Can't use bc each group must have at least two observations.

Asthma_plantcount_lm <- lm(data = join_HealthImpacts_plantcount, EP_ASTHMA ~ plantcount)
summary(Asthma_plantcount_lm)

Asthma_plantcount_plot <- ggplot(join_HealthImpacts_plantcount, 
                          aes(x = plantcount, y = EP_ASTHMA)) +
      geom_point() +
      ggtitle("Test") + 
      labs(x="Number of Plants in Tract",
           y="Percentage of Tract with Asthma (%)")
print(Asthma_plantcount_plot)
```


```{r Relationship between Asthma Prevalence and Health Insurance Status}
##Visualize the relationship between asthma prevalence and health insurance status ##
Asthma_uninsur_plot <- ggplot(intersect_plants_tracts, 
                          aes(x=EP_ASTHMA, y=EP_UNINSUR)) +
      geom_point() +
      geom_smooth(method = lm, color = "black") +
      ggtitle("Relationship between Percentage without Health Insurance and Asthma Prevalence") + 
      labs(x="Percentage of Population with Asthma (%)",
           y="Percentage of Population without Health Insurance (%)")
print(Asthma_uninsur_plot)
```

```{r Relationship between NOx Emissions and SO2 Emissions}

## Might plants those plants that emit high levels of NOx overlap with those that emit high levels of SO2? ## 
NOx_SO2_plot <- ggplot(intersect_plants_tracts, 
                          aes(x=PLNOXAN, y=PLSO2AN)) +
  geom_point() +
  # adjusting axes to hide extreme values
  xlim(0, 800) +
  ylim(0,600) +
  # finding a line of best fit
  geom_smooth(method = lm, color = "black") +
  ggtitle("Relationship between NOx and SO2 Emissions") + 
  labs(x="Plant Annual NOx Emissions (tn)", y="SO2")
print(NOx_SO2_plot)

SO2_NOx_lm <- lm(data = intersect_plants_tracts, PLSO2AN ~ PLNOXAN)
summary(SO2_NOx_lm)
# Residuals range from -491.27 to 1260.12
# Intercept is 0.46503 so there is a positive relationship which means when NOx increase, SO2 tend to increase as well. p-value: < 2.2e-16 which is <.05, therefore it is worthwhile to try to estimate SO2 based on depth information. Although the residual standard error is 238.7 on 199 degrees of freedom, the R-squared value of 0.8274 shows that NOx emissions explain around 82.74% of the variability of SO2 emissions. 
plot(SO2_NOx_lm)
cor.test(intersect_plants_tracts$PLSO2AN, intersect_plants_tracts$PLNOXAN)
# The correlation test returns a correlation strength of 0.9096, which signifies a strong correlation between NOx emissions and SO2 emissions. In summary, there is a strong positive correlation.  
```

```{r Other Exploratory Regression Analyses}

## Visualize the relationship between NOx and Asthma ##
NOx_asthma_plot <- ggplot(intersect_plants_tracts, 
                          aes(x=PLNOXAN, y=EP_ASTHMA,
                              color = PLFUELCT)) +
  geom_point() +
  # adjusting axes to hide extreme values
  xlim(0, 4000) +
  # ylim(0,35) +
  # finding a line of best fit
  geom_smooth(method = lm, color = "black") +
  ggtitle("Prevalence of Asthma  by Depth (meters)") + 
  labs(x="NOx Emissions from Power Plants in Tract (tn)", y="Percentage of Population with Asthma (%)")
print(NOx_asthma_plot)
# unclear relationship when all power plants are included

#Filter for Gas, Biomass, Oil, Coal
ff_intersect_plants_tracts <- intersect_plants_tracts %>%
  filter(PLFUELCT %in% c("GAS", "OIL", "COAL", "BIOMASS"))

## Visualize the relationship between NOx and Asthma  ##
NOx_asthma_ff_plot <- ggplot(ff_intersect_plants_tracts, 
                          aes(x=PLNOXAN, y=EP_ASTHMA,
                              color = PLFUELCT)) +
  geom_point() +
  # adjusting axes to hide extreme values
   xlim(0, 1000) +
  # ylim(0,35) +
  # finding a line of best fit
  geom_smooth(method = lm, color = "black") +
  ggtitle("Prevalence of Asthma in Relation to NOx Emissions") + 
  labs(x="NOx Emissions from Power Plants in Tract (tn)", y="Percentage of Population with Asthma (%)")
print(NOx_asthma_ff_plot)

drop_intersect_plants_tracts <- intersect_plants_tracts %>%
  drop_na(EP_ASTHMA, PLFUELCT, PLNOXAN, PLSO2AN, E_PM, E_DSLPM, E_OZONE)

## Multiple Regression Analysis ##
#Running Akaike's Information Criterion to determine what set of explanatory variables might be best suited to explain asthma prevalence.
Asthma_AIC <- lm(data = drop_intersect_plants_tracts, EP_ASTHMA ~ 
                           PLFUELCT + PLNOXAN + PLSO2AN + 
                           E_PM + E_DSLPM + E_OZONE)
# Choosing a model by AIC in a Stepwise Algorithm
step(Asthma_AIC)
# Starting AIC is -15.48
# From the original, full regression, removing PLSO2AN reduces AIC to -17.4763--it is the most important variable to remove, since a lower AIC is desirable. After removing PLSO2AN to obtain an AIC of -17.48, the next best move would be to remove PLNOXAN, returning an AIC of -19.4554. From there, the last variable to remove would be E_DSLPM, which would further reduce AIC to -19.5518. Removal of each of these variables would make for a better regression than the full, starting regression. Removing E_PM, PLFUELCT, or E_OZONE would make the model worse, as the removal of each increases the AIC from its starting value (increasing AIC to -10.3231, -8.4663, and 7.1690, respectively). Given the 6 starting explanatory variables, the best regression available is to consider PLFUELTCT, E_PM, and E_OZONE. 

#Run regression with the three optimal explanatory variables
Asthma_AIC_opt <- lm(data = drop_intersect_plants_tracts, EP_ASTHMA ~ 
                           PLFUELCT + E_PM + E_OZONE)
summary(Asthma_AIC_opt)
# Multiple R-squared 0.2169 so only 21.69% of the variability on asthma prevalence across tracsts is due to EPM, EOZONE, and PFUEL. Adjusted R-squared 0.1927. p-value is 1.235e-08 so < .05, therefore the relationship among the variable is significantly different than zero. 194 degrees of freedom with residual standard error of 0.9364. Increasing the number of variables in the model should reduce the degrees of freedom.

#Plot relationship between ozone and asthma prevalence
Ozone_asthma_plot <- ggplot(subset(drop_intersect_plants_tracts, COUNTY="Robeson"), 
                          aes(x=E_OZONE, y=EP_ASTHMA)) +
  geom_point() +
  # adjusting axes to hide extreme values
  # xlim(0, 1000) +
  # ylim(0,35) +
  # finding a line of best fit
  geom_smooth(method = lm, color = "black") +
  ggtitle("Prevalence of Asthma") + 
  labs(x="Annual mean days above O3 regulatory standard", y="Percentage of Population with Asthma (%)")
print(Ozone_asthma_plot)

# PLNOXAN ~ Asthma
NOx_asthma_lm <- lm(EP_ASTHMA ~ PLNOXAN, data = intersect_plants_tracts)
summary(NOx_asthma_lm)
# p-value < .05 (p-value = .004137) 
# we can reject the null hypothesis that there is no statistically significant relationship between asthma prevalence and NOx emitted from power plants
# 199 degrees of freedom
# residual standard error of 1.023
cor.test(intersect_plants_tracts$EP_ASTHMA, intersect_plants_tracts$PLNOXAN)
# -0.20 signifies a negative correlation, but 
# a weak correlation between the two variables
plot(NOx_asthma_lm)

namepcap_healtheff <- lm(Sum_NAMEPCAP ~ EP_BPHIGH + EP_ASTHMA + E_PM + E_DSLPM + E_OZONE, data = EBM_nameplate_merge)
step(namepcap_healtheff)
#Keep only EP_ASTHMA (AIC = 34593) because all other variables increase AIC
knitr::kable(summary(namepcap_healtheff)$coef, format = "html", caption = "Linear Regression Results - Nameplate Capacity and Air Quality")

namepcap_asthma <- lm(data = EBM_nameplate_merge, Sum_NAMEPCAP ~ EP_ASTHMA)
summary(namepcap_asthma)
plot(namepcap_asthma)

#Run simple regression
capacity_income_model <- lm(NAMEPCAP_plant ~ MEDHHINC_2021, data = merged_data_capacity)

knitr::kable(summary(capacity_income_model)$coef, format = "html", caption = "Linear Regression Results - Namepate Capacity")

#To what extent does high PM exposure explain asthma in NC 
PM_asthma_lm<- lm(data = EBM_countytract, EP_ASTHMA ~ E_PM)
summary(PM_asthma_lm)
# p-value < 0.05
# the R-squared value of .0097 signifies that elevated PM concentrations only explain less than 1% of the variability in asthma in NC tracts
cor.test(EBM_countytract$EP_ASTHMA, EBM_countytract$E_PM)
# -.098 signifies a negative and significant but not super strong correlation

#To what extent does high PM exposure explain high blood pressure in NC 
PM_bphigh_lm<- lm(data = EBM_countytract, EP_BPHIGH ~ E_PM)
summary(PM_bphigh_lm)
# p-value < 0.05
# the R-squared value of .0927 signifies that elevated PM concentrations explain 9.27% of the variability in high blood pressure in NC tracts
cor.test(EBM_countytract$EP_BPHIGH, EBM_countytract$E_PM)
# -0.30 signifies a negative and significant but not super strong correlation

#Multiple regression looking at whether coal proximity, PM, and ozone might explain high blood pressure
EBM_highbp_lm <- lm(EP_BPHIGH ~ E_PM + E_DSLPM + E_OZONE, data=EBM_countytract)
step(EBM_highbp_lm)

#Multiple regression looking at whether coal proximity, PM, and ozone might explain asthma
EBM_asthma_lm <- lm(EP_ASTHMA ~ E_PM + E_DSLPM + E_OZONE, data=EBM_countytract)
step(EBM_asthma_lm)

#Run the multiple regression to determine whether there is a relationship
EBM_multivar_lm <- lm(EP_UNINSUR ~ EP_ASTHMA + EP_BPHIGH + E_IMPWTR + 
                        E_COAL + E_PM + E_OZONE,
                        data=EBM_countytract)
step(EBM_multivar_lm)

EBM_lead_wtr_lm <- lm(EP_UNINSUR ~ E_IMPWTR + 
                        E_PM,
                        data=EBM_countytract)
step(EBM_lead_wtr_lm)
#Summarize the multiple regression
summary(EBM_multivar_lm)
```
NOxSO2 Analyses

```{r Relationship between HV and emissions, include=FALSE}

HV_emissions_multivar <- lm(data = nona_sums_intersection_tract, RPL_HVM ~ Tract_PLNOXAN + Tract_PLSO2AN + Tract_PLCH4AN + Tract_UNNOX + Tract_UNSO2 + Tract_UNCH4)
step(HV_emissions_multivar)
# Residual standard error: 0.298 on 38 degrees of freedom
#   (269 observations deleted due to missingness)
# Multiple R-squared:  0.04781,	Adjusted R-squared:  -0.05242 
# F-statistic: 0.477 on 4 and 38 DF,  p-value: 0.7524
# The AIC stepwise algorithm resulted in removal being recommended for all variables. None of the variables were found to have an explanatory relationship worth exploring for the overall health vulnerability module (starting AIC was -98.35 and removal of each variable lowered AIC).
```

The following section of analyses explore relationships between emissions of SO2 and NOx and respiratory effects using eGRID data as well as data from the Environmental Burden and Health Vulnerability modules in the Environmental Justice Index (EJI). The development of the EJI is relatively new and the index has limitations. In accordance with its limitations, which can be found in detail within the technical documentation linked in the Metadata folder, this high-level exploration is not intended to define individuals at risk, represent risk or exposure for a community, or suggest that the markers explored form a full or near-full representation of environmental or health characteristics. This brief exploration merely examines whether a statistically significant relationship can be detected among select local factors such as power plant emissions and human health effects at a point in time across North Carolina. Based on established knowledge regarding air pollution and its impacts to human health, the expectation would be that higher emissions of NOx and SO2 are associated with higher rates of respiratory disease. 


```{r NOx SO2 Correlation Tests, include=FALSE}

Tract_UNNOX_UNSO2_lm <- lm(data = sums_intersection_tractlevel, Tract_UNNOX ~ Tract_UNSO2)
summary(Tract_UNNOX_UNSO2_lm)
# Residual standard error: 905 on 41 degrees of freedom
#   (269 observations deleted due to missingness)
# Multiple R-squared:  0.324,	Adjusted R-squared:  0.3075 
# F-statistic: 19.65 on 1 and 41 DF,  p-value: 6.809e-05

# STRONGEST CORRELATION R-SQUARED AND LOWEST P-VALUE -- graph this one!
Tract_PLNOXAN_PLSO2AN_lm <- lm(data = sums_intersection_tractlevel, Tract_PLNOXAN ~ Tract_PLSO2AN)
summary(Tract_PLNOXAN_PLSO2AN_lm)
# Residual standard error: 368.2 on 41 degrees of freedom
#   (269 observations deleted due to missingness)
# Multiple R-squared:  0.8889,	Adjusted R-squared:  0.8862 
# F-statistic:   328 on 1 and 41 DF,  p-value: < 2.2e-16

Plant_UNNOX_UNSO2_lm <- lm(data = intersection_plants_EJI, UNNOX ~ UNSO2)
summary(Plant_UNNOX_UNSO2_lm)
# Residual standard error: 779 on 70 degrees of freedom
#   (468 observations deleted due to missingness)
# Multiple R-squared:  0.3756,	Adjusted R-squared:  0.3667 
# F-statistic: 42.11 on 1 and 70 DF,  p-value: 1.051e-08

# STRONGEST CORRELATION R-SQUARED AND LOWEST P-VALUE -- graph this one!
Plant_PLNOXAN_PLSO2AN_lm <- lm(data = intersection_plants_EJI, PLNOXAN ~ PLSO2AN)
summary(Plant_PLNOXAN_PLSO2AN_lm)
# Residual standard error: 421.2 on 70 degrees of freedom
#   (468 observations deleted due to missingness)
# Multiple R-squared:  0.8114,	Adjusted R-squared:  0.8087 
# F-statistic: 301.2 on 1 and 70 DF,  p-value: < 2.2e-16

```

```{r Relationship between Asthma and NOx and SO2, include=FALSE}

Asthma_emissions_multivar <- lm(data = nona_sums_intersection_tract,
                            EP_ASTHMA ~ Tract_PLNOXAN + Tract_PLSO2AN + Tract_PLCH4AN)
step(Asthma_emissions_multivar)
# Optimized based on stepwise regression
Asthma_emissions_opt <- lm(data = nona_sums_intersection_tract,
                            EP_ASTHMA ~ Tract_PLCH4AN)
summary(Asthma_emissions_opt)
# Residual standard error: 1.045 on 41 degrees of freedom
# Multiple R-squared:  0.0528,	Adjusted R-squared:  0.0297 
# F-statistic: 2.286 on 1 and 41 DF,  p-value: 0.1382

Asthma_NOxSO2_multivar <- lm(data = nona_sums_intersection_tract,
                            EP_ASTHMA ~ Tract_PLNOXAN + Tract_PLSO2AN)
step(Asthma_NOxSO2_multivar)
summary(Asthma_NOxSO2_multivar)
#Residual standard error: 1.065 on 40 degrees of freedom
# Multiple R-squared:  0.03935,	Adjusted R-squared:  -0.008688 
# F-statistic: 0.8191 on 2 and 40 DF,  p-value: 0.4481
```

Knowing that NOx contribute to atmospheric ozone formation in addition to having direct human health effects, we were curious whether a multiple regression of annual mean days above O3 regulatory standard (E_OZONE) against emissions of NOx, SO2, CO2, and CH4 might elevate NOx emissions from power plants as the variable that might most explain E_OZONE. While it performed as the strongest amongst the multiple regression variables, it ultimately did not return a significant relationship when run as a single linear regression against E_OZONE (the p-value of the single linear regression was 0.985). While a relationship between the two variables could not be concluded and a lack of relationship could not be rejected, it is worth noting that E_OZONE is is annual mean days above O3 regulatory standard calculated using a 3-year average--it is not the atmospheric concentration level measured at each tract.

```{r Relationship between atmospheric ozone levels and emissions, include=FALSE}

Ozone_emissions_multivar <- lm(data = nona_sums_intersection_tract,
                            E_OZONE ~ Tract_PLNOXAN + Tract_PLSO2AN + Tract_PLCH4AN)
step(Ozone_emissions_multivar)
# Start:  AIC=-65.9
# E_OZONE ~ Tract_PLNOXAN + Tract_PLSO2AN + Tract_PLCH4AN
# 
#                 Df Sum of Sq    RSS     AIC
# - Tract_PLCH4AN  1   0.23392 7.9446 -66.615
# - Tract_PLSO2AN  1   0.25122 7.9619 -66.521
# <none>                       7.7107 -65.900
# - Tract_PLNOXAN  1   1.13494 8.8456 -61.995
# 
# Step:  AIC=-66.61
# E_OZONE ~ Tract_PLNOXAN + Tract_PLSO2AN
# 
#                 Df Sum of Sq    RSS     AIC
# <none>                       7.9446 -66.615
# - Tract_PLNOXAN  1    1.0030 8.9476 -63.502
# - Tract_PLSO2AN  1    1.0338 8.9784 -63.354
# 
# Call:
# lm(formula = E_OZONE ~ Tract_PLNOXAN + Tract_PLSO2AN, data = nona_sums_intersection_tract)
# 
# Coefficients:
#   (Intercept)  Tract_PLNOXAN  Tract_PLSO2AN  
#     0.7098862      0.0004248     -0.0010003  

opt_Ozone_emissions_multivar <- lm(E_OZONE ~ Tract_PLNOXAN + Tract_PLSO2AN, data = nona_sums_intersection_tract)
summary(opt_Ozone_emissions_multivar)
# Residual standard error: 0.4457 on 40 degrees of freedom
# Multiple R-squared:  0.1168,	Adjusted R-squared:  0.07261 
# F-statistic: 2.644 on 2 and 40 DF,  p-value: 0.08345
```

##### Figure 25: The relationship between adjusted NOx emissions and SO2 emissions from North Carolina power plants in 2021. As noted in the eGRID metadata file, certain plants' emissions may be adjusted if the plant is a CHP (combined heat and power) facility or if one or more of its units burns biomass. {-}

```{r Plotting relationship between adjusted NOx emissions and SO2 emissions, echo=FALSE}

## Might plants those plants that emit high levels of NOx overlap with those that emit high levels of SO2? ## 
NOx_SO2_plot <- ggplot(nona_sums_intersection_tract, 
                          aes(x=Tract_PLNOXAN, y=Tract_PLSO2AN)) +
  geom_point() +
  # adjusting axes to hide extreme values
  xlim(0, 500) +
  ylim(0,500) +
  # finding a line of best fit
  geom_smooth(method = lm, color = "black") +
  ggtitle("Relationship between Adjusted NOx and SO2 Emissions by Tract") + 
  labs(x="Tract Annual NOx Emissions \n (short tons)", y="Tract Annual SO2 Emissions \n (short tons)")
print(NOx_SO2_plot)
```

##### Figure 26: The relationship between unadjusted NOx emissions and SO2 emissions from North Carolina power plants in 2021. {-}

```{r Plotting relationship between unadjusted NOx SO2 emissions, echo=FALSE}


## Might plants those plants that emit high levels of NOx overlap with those that emit high levels of SO2? ## 
UNNOx_UNSO2_plot <- ggplot(intersection_plants_EJI, 
                          aes(x=PLNOXAN, y=PLSO2AN)) +
  geom_point() +
  # adjusting axes to hide extreme values
  xlim(0, 1250) +
  ylim(0, 2000) +
  # finding a line of best fit
  geom_smooth(method = lm, color = "black") +
  ggtitle("Relationship between Adjusted NOx and SO2 Emissions by Plant") + 
  labs(x="Plant Annual NOx Emissions \n (short tons)", y="Plant Annual SO2 Emissions \n (short tons)")
print(UNNOx_UNSO2_plot)
```

Given that NOx (including NO2) and SO2 both have negative human health impacts and especially respiratory impacts, we were curious to explore whether those plants that emit high levels of NOx might overlap with those that emit high levels of SO2. Plotting the relationship as in Figures 25 and 26 above appears to show a relationship between emitted NOx and SO2 from power plants in 2021, however, the relationship merits further exploration through linear regression and correlation testing. If power plants that are emitting high levels of NOx are also emitting high levels of SO2, this could be impactful information for air quality management professionals looking to make decisions on how to reduce emissions from power plant sources and which plants to prioritize. 

```{r Testing correlations between NOx and SO2, include=FALSE}

# eGRID emissions data had many NAs
# Removing NAs for multivariate regression analyses
nona_intersection_plants_EJI <- intersection_plants_EJI %>%
  select(GEOID, CountyFIPS, StateAbbr, COUNTY, E_TOTPOP, M_TOTPOP, 
          PLNGENAN, NAMEPCAP, PLNOXAN, PLSO2AN,
          PLCH4AN, UNNOX, UNSO2, UNCH4, RPL_EBM, 
          RPL_HVM, RPL_SVM, E_PM, E_OZONE, E_DSLPM, E_COAL, EP_UNINSUR, 
          EP_AGE65, EP_ASTHMA, EP_BPHIGH, EP_CANCER, EP_MHLTH, geometry) %>%
  drop_na()

SO2_NOx_lm <- lm(data = nona_intersection_plants_EJI, PLSO2AN ~ PLNOXAN)
summary(SO2_NOx_lm)
# Residual standard error: 221 on 70 degrees of freedom
# Multiple R-squared:  0.8114,	Adjusted R-squared:  0.8087 
# F-statistic: 301.2 on 1 and 70 DF,  p-value: < 2.2e-16
##################### !!!!!!!!!!!!!!!!
# Residuals range from -491.27 to 1260.12
# Intercept is 0.46503 so there is a positive relationship which means when NOx increase, SO2 tend to increase as well. p-value: < 2.2e-16 which is <.05, therefore it is worthwhile to try to estimate SO2 based on depth information. Although the residual standard error is 238.7 on 199 degrees of freedom, the R-squared value of 0.8274 shows that NOx emissions explain around 82.74% of the variability of SO2 emissions. 
# plot(SO2_NOx_lm)

# Pearson's - cor .90
cor.test(nona_intersection_plants_EJI$PLSO2AN, nona_intersection_plants_EJI$PLNOXAN)

# Kendall tau - tau .45 and z = 5.5608, p-value = 2.685e-08
#alternative hypothesis: true tau is not equal to 0
cor.test(nona_intersection_plants_EJI$PLSO2AN, nona_intersection_plants_EJI$PLNOXAN, method="kendall")

# Spearman
# S = 30268, p-value = 4.003e-06
# alternative hypothesis: true rho is not equal to 0
# sample estimates:
#       rho 
# 0.5133525 
cor.test(nona_intersection_plants_EJI$PLSO2AN, nona_intersection_plants_EJI$PLNOXAN, method="spearman")
# The correlation test returns a correlation strength of 0.9096, which signifies a strong correlation between NOx emissions and SO2 emissions. In summary, there is a strong positive correlation.

cor.test(nona_sums_intersection_tract$EP_ASTHMA, nona_sums_intersection_tract$Tract_Generation, method = "kendall")
# z = -2.0226, p-value = 0.04312
# alternative hypothesis: true tau is not equal to 0
# sample estimates:
#       tau 
# -0.216139 


cor.test(nona_sums_intersection_tract$EP_ASTHMA, nona_sums_intersection_tract$EP_UNINSUR, method = "kendall")
# t = 3.4586, df = 41, p-value = 0.00128
# alternative hypothesis: true correlation is not equal to 0
# 95 percent confidence interval:
#  0.2040226 0.6787116
# sample estimates:
#       cor 
# 0.4752449 
# Kendall: data:  nona_sums_intersection_tract$EP_ASTHMA and nona_sums_intersection_tract$EP_UNINSUR
# z = 2.966, p-value = 0.003017
# alternative hypothesis: true tau is not equal to 0
# sample estimates:
#       tau 
# 0.3172807 

cor.test(nona_sums_intersection_tract$EP_BPHIGH, nona_sums_intersection_tract$EP_UNINSUR)
# t = 3.1927, df = 41, p-value = 0.002707
# alternative hypothesis: true correlation is not equal to 0
# 95 percent confidence interval:
#  0.1684516 0.6583340
# sample estimates:
#       cor 
# 0.4462197 

```

End
\